---
title: "3. Grid construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Grid construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(icosa)
library(rgl)
library(knitr)
knit_hooks$set(rgl = hook_rgl)

knitr::knit_hooks$set(
  source = function(x, options) {
    hook.r = function(x, options) {
      fence <- "```"
      language = tolower(options$engine)
      if (language == 'node') language = 'javascript'
      if (!options$highlight) language = 'text'
      if(!is.null(options$foldcode)) {
      paste0('\n\n', "<details><summary>Source</summary>\n", fence, language, '\n', x, fence,  '\n\n', "</details>\n")
      } else {
              paste0('\n\n', fence, language, '\n', x, fence,  '\n\n')
      }
    }
    x = knitr:::hilight_source(x, 'markdown', options)
    hook.r(
      paste(c(
        x, 
        ''
      ), collapse = '\n'), 
      options
    )
  }
)
```

# Introduction

Grid objects are used as reference points for the geometry of 2D and 3D operations. This tutorial focuses on the grid structures, how the grids are derived, how and the types that can be created with `icosa`. 

## Grids as objects

In their basic form, the grids themselves are 3D objects, that consist of **vertices** (points), **edges** (segments between two points), and **faces** (polygons defined with a sequence of points). In spatial analyses, faces are usually referred to as **cells**. 

Note that only points have 3D information, both edges and faces are defined using them: which effectively means that that it is up to the interpretation whether edges are interpreted as linear segments (that define polyhedra), or arcs: i.e. portions of great circle segments. Similarly, faces can be interpreted both as planar triangles and as spherical triangles. All calculations in `icosa` are implemented in XYZ Cartesian space. 

In order to derive the grids `icosa` uses a spherical model for Earth, which is based on the [authalic radius]() (6371.007 km, the radius of a sphere that has the same volume as the actual shape of the planet) as defined in [ESRI:37008](). 

## The ideal grid

An optimal horizontal grid would have completely uniform structure, with its points distributed completely evenly on the sphere. This means that faces (i.e. cells) defined using these points would have identical **size**, **shape** and **relative distance** from each other - a criterion that that turns out to be problematic: only the platonic solids have vertices that fulfill these criteria perfectly. 


In the `icosa` package all grids are derived from the [regular icosahedron](), a platonic solid that has the highest number of faces, which is 20 (seen below):

```{r plot3d, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
library(icosa)
library(rgl)

# create a trigrid class object
tri <- trigrid()

# plot the object in 3d
# warnings occur on some systems due to rgl fonts
suppressWarnings(faces3d(tri, guides=F, col="gray"))

# rotate viewport
view3d( theta =30, phi = 20)

```

This coarse resolution  is hardly acceptable for the analysis of spatial heterogeneity, which means that we have to find a way to increase the number of cells in a uniform way. 

## Tessellation

Tessellation is the process of taking a polygon (i.e. triangle), and breaking it up into smaller triangles. Tessellation of a regular triangle is trivial, but it is useful to illustrate the concept (e.g. below). 


```{r tessel, fig.width=10, fig.height=4, foldcode=TRUE}
# very verbose code to produce the figure below...
par(mfrow=c(1,3))

# This is a beautiful programming exercise:
# can you implement a function that does this for an arbitrary number of edges?

# tessellation = 1
plot(
	NULL, NULL, xlim=c(0,1), ylim=c(0,1), axes=FALSE, 
	xlab="Tessellation (#new edge on triangle side) = 1", 
	ylab="", asp=1, cex.lab=1)

# original points of a regular triangle
A <- c(0,0)
B <- c(1,0)
C <- c(0.5, sqrt(3)/2)

# triangle- as matrix
triangle <- rbind(A, B, C)

# draw
polygon(triangle, col=rgb(0.34,0.53,0.2, 0.5))

# the point
text(
	x=triangle[,1], y=triangle[,2], 
	label=rownames(triangle), pos=c(1,1, 3))

# ------------------------------------------------------------------------------
# tessellation =2
# breaking segments in two - mid between points
ab <- (A+B)/2
bc <- (B+C)/2
ca <- (C+A)/2

# create new set of triangles
triangle1 <- rbind(A, ab, ca)
triangle2 <- rbind(ca, ab, bc)
triangle3 <- rbind(ab, B, bc)
triangle4 <- rbind(ca, bc, C)

# empty plot
plot(
	NULL, NULL, xlim=c(0,1), ylim=c(0,1), axes=FALSE, 
	xlab="Tessellation (#new edge on triangle side) = 2", 
	ylab="", asp=1, cex.lab=1)

# te colors of the triangles
greens <- rgb(0.34,0.53,0.2, seq(0,1,length.out=4+2))

# plot the triangles
polygon(triangle1, col=greens[2])
polygon(triangle2, col=greens[3])
polygon(triangle3, col=greens[4])
polygon(triangle4, col=greens[5])

# the points
unpoint <- rbind(A, B, C, ab, bc, ca)

# putl labels
text(
	x=unpoint[,1], y=unpoint[,2], 
	label=rownames(unpoint), pos=c(1,1, 3, 1, 4, 2))

# ------------------------------------------------------------------------------
# tessellation = 3

# points derived with simple coordinate geometry
# difference vector reduced, offset from subtracted
a1b <- (B-A)/3 + A
a2b <- (B-A)/3*2 + A
b1c <- (C-B)/3 + B
b2c <- (C-B)/3*2 + B
c1a <- (C-A)/3 + A
c2a <- (C-A)/3*2 + A
mid <- (A+B+C)/3

# the triangles
triangles <- list()
triangles[[1]] <- rbind(A, a1b, c1a)
triangles[[2]] <- rbind(a1b, a2b, mid)
triangles[[3]] <- rbind(a2b, B, b1c)
triangles[[4]] <- rbind(a1b, mid, c1a)
triangles[[5]] <- rbind(a2b, b1c, mid)
triangles[[6]] <- rbind(c1a, mid, c2a)
triangles[[7]] <- rbind(mid, b1c, b2c)
triangles[[8]] <- rbind(mid, b2c, c2a)
triangles[[9]] <- rbind(c2a, b2c, C)

# define color sequence
greens <- rgb(0.34,0.53,0.2, seq(0,1,length.out=length(triangles)+2))

# empty plot
plot(
	NULL, NULL, xlim=c(0,1), ylim=c(0,1), axes=FALSE,
	xlab="Tessellation (#new edge on triangle side) = 3",
	 ylab="", asp=1, cex.lab=1)

# plotting the triangles
for(i in 1:length(triangles)) polygon(triangles[[i]], col=greens[i+1])

# the point coordinates 
unpoint <- rbind(
	A, B, C, 
	a1b, a2b, b1c, b2c, c1a, c2a,
	mid)

# the names of the points
text(
	x=unpoint[,1], y=unpoint[,2], 
	label=rownames(unpoint),
	pos= c(1, 1, 3, 1, 1, 4, 4, 2, 2, 1))

```

In this case, the edges of the triangles are split to equal segments, which in turn define smaller regular triangles. The number of resulting segments on an original edge defines the resulting number of triangles: if there are *n* new segments, there will be $n^2$ number of triangles. 

Moreover, once the new triangles are created, the process can be repeated recursively, creating a newer, smaller set of triangles. This recursive iteration can be represented as a vector of positive integer numbers (which we can call the **tessellation vector**), for example:

```R
c(2,3,2)
```

Which indicates that that the edge of a triangle is to split once into two (leading to 4 triangles), and then all resulting edges to be split into 3 (creating $4 * 3^2 = 36$ triangles), and then the resulting edges are also split into 2, leading to a total of $36 * 4 = 144$ triangles. This means that a total of $2 * 3 * 2 = 12$ (i.e. the product of the tessellation vector) resulting edges will be created for every original edge (note that $12^2 = 144$!). In the planar case and with regular triangles, this leads to the same result as doing a single 12-fold subdivision, but leads to slightly different results in the spherical case. 

Since an icosahedron has 20 triangles/faces, the total number of faces will be $n^2 * 20$. Therefore the **tessellation vector** is the control of grid resolution.

Despite its intuitive nature, the implementation of this process is not trivial for spherical triangles (which have arcs instead of linear segments as edges) and becomes a source of distortions ([see below]()). The exact implementation will be explained in a later article focusing on the tessellation. 

* * *

# Triangular grids


Triangular grids are directly dervied from the regular icosahedron using a spherical version of the tessellation method described above. These (and all other grids, including the [`hexagrid`s]()) belong in the `trigrid-class` and can be constructed with the `trigrid()` function. 

## A primitive triangular grid 

Without any arguments, this will construct a regular icosahedron - with all the attributes that any other grid object has. 

```{r trifirst}
# loading the package
library(icosa)

# create a trigrid class object
tri <- trigrid()

# the show() method displays basic information of the object 
tri
```

All calculations are natively done in xyz Cartesian space, and (as indicated above) this can be visualized with 3D plotting tools. The implemented 3d plotting functions work with R's interactive implementation of openGL (in the CRAN package `rgl`) - see later tutorials about details.

A simple icosahedron looks as such (with an internal sphere added to make the render more intuitive to understand): 

```{r plot3dtri, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
# warnings occur on some systems due to rgl fonts
suppressWarnings(plot3d(tri, guides=F))
```

Without any specified additional entry, the created object will an icosahedron with the center of ``c(0,0,0)`` Cartesian coordinates and the authalic radius of Earth used as the distance between the object center and the vertices. These can be altered by setting the ``radius`` and ``center`` arguments if necessary (e.g. for composing 3d imagery). 

## Tessellating a triangular grid

Setting the first, `tessellation` argument  of the ``trigrid()`` function can be used to create higher-resolution grids. The input to this argument is a **tessellation vector** as described [above](). For instance this line of code will instantiate a triangular grid with the tessellation vector set to `8` (every side of the icosahedron is split into 8 new edges):

```{r trigrid_create}
# create a trigrid class object
gl <- trigrid(tessellation=c(8))
gl
```

```{r trigrid_3d, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
# plot the object in 3d
suppressWarnings(plot3d(gl, guides=F))
```

Choosing a vector with a higher product leads to a more complex, higher-resolution grid:

```{r trigrid_higher}
# create a trigrid class object
gh <- trigrid(tessellation=c(4,4))
gh
```

```{r trigrid_higher_3d, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
# plot the object in 3d
suppressWarnings(plot3d(gh, guides=F))
```

## Controlling grid resolution

You have probably noted the decrease in overall performance - which is the overhead of the 3D implementation and the flexibilty that these grids offer: the higher the tessellation, the longer it takes to calculate a grid, and the bigger it is in the computer's memory. Also, tessellation is easy to understand from the point of creating a grid, but it is very unintiutive from the point of application, where we tend to think in *degrees* and *kilometers*. If you would like to explore the connection between grid resolution, size, performance and tesselation, you can do so using the `triguide` built-in object.

```{r triguide}
data(triguide)
str(triguide)
```

Check out `?triguide` for the meaning of the columns
You can use this object to  

To make the instantiation of grids somewhat more straightforward, you can also provide the desired edge-length in degrees (argument `deg`), and tessellation vector that leads to the closest result will be provided to you. For instance, if you would like to have a grid with approximately 5° of edge length, you can get it with:

```{r trideg}
five <- trigrid(deg=5)
five
```

Note that this will not lead to an exactly 5°- resolution grid, but it still can serve as a good approximation. 

# Hexagonal grids

## Rationale

Triangular grids are intutive ways of breaking up a spherical surface to similar entities and they can be used for discretization of spatial. You could register data at the centers of the triangles, but there are several problems with that: triangular cells are very anisometric, the vertices are very far away from the cell centerpoints compared to the middle of the edges, so the degree to which a single point can represent points in the cell varies. 

```{r cells, fig.width=10, fig.height=4, foldcode=TRUE}
par(mfrow=c(1,4))

# This is a beautiful programming exercise:
# can you implement a function that does this for an arbitrary number of edges?
green <- rgb(0.34,0.53,0.2, 0.5)

# triangle
plot(
	NULL, NULL, xlim=c(0,1), ylim=c(0,1), axes=FALSE, 
	xlab="", 
	ylab="", asp=1, cex.lab=1)

# original points of a regular triangle
A <- c(0,0)
B <- c(1,0)
C <- c(0.5, sqrt(3)/2)

# triangle- as matrix
triangle <- rbind(A, B, C)

# draw
polygon(triangle, col=green)
text(triangle[,1],triangle[,2], label=rownames(triangle),pos=c(1,1,3))

# triangle midpoint
mid <- (A+B+C)/3
text(label="O", x=mid[1], y=mid[2], pos=1)

# midpoint - vertex
segments(x0=mid[1], x1=B[1], y0=mid[2], y1=B[2])
midb <- (B+mid)/2
text(x=midb[1], y=midb[2], label=paste0("a = ", round(sqrt(sum((mid-midb)^2)),2)), pos=4)


# midpoint - edgemid
bc <- (B+C)/2
segments(x0=mid[1], x1=bc[1], y0=mid[2], y1=bc[2], col="gray50")
midbc <- (mid+bc)/2
text(x=midbc[1], y=midbc[2], label=paste0("b = ", round(sqrt(sum((mid-midbc)^2)),2)), pos=2, col="gray50")

# ------------------------------------------------------------------------------
# square

plot(
	NULL, NULL, xlim=c(-0.05,1.05), ylim=c(-0.05,1.05), axes=FALSE, 
	xlab="", 
	ylab="", asp=1, cex.lab=1)

A <- c(0,0)
B <- c(1,0)
C <- c(1,1)
D <- c(0,1)

square <- rbind(A, B, C, D)
polygon(square, col=green)

text(square[,1],square[,2], label=rownames(square),pos=c(1,1,3,3))

# midpoint
mid <- (A + B + C + D)/4
text(x=mid[1], y=mid[2], label="O", pos=1)

# mid-vertex

segments(x0=mid[1], x1=1, y0=mid[2], y1=1)


# mid edge mid
segments(x0=mid[1], x1=0.5, y0=mid[2], y1=1)


# ------------------------------------------------------------------------------
# hexagon

plot(
	NULL, NULL, xlim=c(-0.05,1.05), ylim=c(-0.05,1.05), axes=FALSE, 
	xlab="", 
	ylab="", asp=1, cex.lab=1)

# ------------------------------------------------------------------------------
# A circle
plot(
	NULL, NULL, xlim=c(0,1), ylim=c(0,1), axes=FALSE, 
	xlab="", 
	ylab="", asp=1, cex.lab=1)

rad <- seq(0, 2*pi, length.out=360)
circle <- cbind(x=cos(rad)*0.5+0.5, y=sin(rad)*0.5+0.5)
polygon(circle, col=green)

text(label="O", x=0.5, y=0.5, pos=1)
O <- c(0.5, 0.5)
A <- circle[45, ]
B <- circle[120, ]

segments(x0=0.5, x1=c(A[1], B[1]), y0=0.5, y1=c(A[2], B[2]))

oa <- (O+A)/2
ob <- (O+B)/2
text(oa[1], oa[2],  pos=4, label="r = 0.5")
text(ob[1], ob[2],  pos=2, label="r = 0.5")
```

Also, some spatial calculations rely on the concept of neighborhood - knowning what cells are in contact. Triangular grids summer from the same issues as rectangular grids: neighborhood can be defined differently, depending on the pattern of contact. Based on analogies from chess, you talk either about a 'rook' or 'queen'-neighborhood, when shared edges or shared vertices are considered to be the base criteria for neighborhood, respectively. 

## Grid inversion

Using hexagons solve both of these problems. Hexagonal-cells can be defined with inversion: replacing the grid vertices with face-centers of the triangular grids, and inserting vertices at the center of of the faces. 
This inversion operation follows the tessellation procedure, and uses already created `trigrid-class` objects and produces `hexagrid-class` entities:

Executing this proces on the icosahedron (`tessellation=1`), ironically, produces an object without any hexagonal faces, the result of this operation is a regular pentagon-dodecahedron: 

```{r hex}
# create a hexagrid class object
dodecahedron <- hexagrid(tessellation=1)
dodecahedron
```

```{r plot3dhex, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
# plot the object in 3d
# warnings occur on some systems due to rgl fonts
suppressWarnings(faces3d(dodecahedron, guides=F, col="gray"))

# rotate viewport
view3d( theta =80, phi = 40)
```

Hexagons are introduced when more complex triangular grids are inverted. 

```{r hex2}
# create a hexagrid class object
hexa <- hexagrid(tessellation=4)
hexa
```

```{r plot3dhex2, rgl=TRUE,fig.width=10, fig.height=10, foldcode=TRUE}
# plot the object in 3d
# warnings occur on some systems due to rgl fonts
suppressWarnings(plot3d(hexa, guides=F))

# rotate viewport
view3d( theta =80, phi = 40)
```

As a matter of fact, it is impossible to create these hexagonal grids without the introduction of pentagons, exactly 12 of them, that are positioned at the vertices of the icosahedron. For this reason, it is better to call these penta-hexagonal grids. 

These objects are practically Goldberg polyhedra. 


## Hexagrid resolution control 

# 2D mapping

# Distortion patterns
