---
title: "1. Installation"
description: >
  Alternatives to install the icosa R package. 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Stable version from the CRAN

The stable version of the package is available on [CRAN servers](https://cran.r-project.org/package=icosa), which means that you can install it with the regular `install.packages()` function.

```{r cran, eval=FALSE}
install.packages("icosa")
```

If for some reason, the installlation from CRAN fails, you might need to look into alternative methods.


* * * 

## Alternative methods

#### Binaries for Windows

To make installation less problematic, the windows binary version of the package are also available on GitHub. This should work with all versions of windows

```{r stablewin, eval=FALSE}
install.packages(
	"https://github.com/adamkocsis/icosa/raw/master/_archive/binaries/Win_x64_x86/icosa_0.11.0.zip", 
	repos=NULL)
```

#### Stable version from GitHub from source

For the sake of completeness and simplicity, the sources of the stable version are also available on GitHub. The `install.packages()` function can be used to install this from source - but you will have to install dependencies of the package, and you have to have a compiler installed on your system 

```{r stable, eval=FALSE}
install.packages(
	"https://github.com/adamkocsis/icosa/raw/master/_archive/source/icosa_0.11.0.tar.gz", 
	repos=NULL, type="source")
```

#### Development version - the raw 

If you encounter technical problems with the package that are not described anywhere, you might want to take a look at the [development version](https://github.com/adamkocsis/icosa/tree/devel).

If you wish to install the development version, I recommend a manual installation:  
1. Clone the repository to your local hard drive.  
2. Make sure that you have the package dependencies, and a compiler installed: on Windows, [RTools](https://cran.r-project.org/bin/windows/Rtools/) should include all the tools you need, on Mac, you will need to have [XCode](https://developer.apple.com/xcode/) - if you use GNU/Linux, you have to go down deeper the rabbit hole :).   
3. Open a terminal and navigate to the directory where you cloned. The `icosa` directory should be visible from there.  
4. Run this line in the terminal  

```
R CMD INSTALL icosa
```

- *If you see an error suggesting that `R` is not found, you have to add it to your `PATH` environmental variable.*  
- *If the R packages that `icosa` depend on are not installed, you have to install them manually, or you will get an error.*
- You can also try `devtools::install_github()`. 
